---
title: "Weather events with most impact on economy and public health in the United States"
author: "Aditya Ishak"
date: "18 November 2015"
output: html_document
keep_md : true
---

# Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States between 1950 and November 2011, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

This report will show which types of events are:

1. The most harmful to population health
2. Have the greatest economic consequences

# Data Processing

Read raw data from file. The data file is available from [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) if you wish to download
```{r cache=TRUE}
# read from the file
rawdata <- read.csv("repdata-data-StormData.csv.bz2")
```

Only select columns that are relevant for the analysis
```{r}
# retrieve columns that we are interested in
tidydata <- rawdata[,c('EVTYPE','FATALITIES','INJURIES', 'CROPDMG', 'PROPDMGEXP', 'PROPDMG', 'CROPDMGEXP')]
```

Perform additional processing on the selected columns
```{r}
# Calculates nominal property damage and store in new column called PROPDMG_NOM
tidydata$PROPDMG_NOM = 0
tidydata[tidydata$PROPDMGEXP == "H", ]$PROPDMG_NOM = tidydata[tidydata$PROPDMGEXP == "H", ]$PROPDMG * 10^2
tidydata[tidydata$PROPDMGEXP == "K", ]$PROPDMG_NOM = tidydata[tidydata$PROPDMGEXP == "K", ]$PROPDMG * 10^3
tidydata[tidydata$PROPDMGEXP == "M", ]$PROPDMG_NOM = tidydata[tidydata$PROPDMGEXP == "M", ]$PROPDMG * 10^6
tidydata[tidydata$PROPDMGEXP == "B", ]$PROPDMG_NOM = tidydata[tidydata$PROPDMGEXP == "B", ]$PROPDMG * 10^9

# Calculate nominal crop damage and store ina new column called CROPDMG_NOM
tidydata$CROPDMG_NOM = 0
tidydata[tidydata$CROPDMGEXP == "H", ]$CROPDMG_NOM = tidydata[tidydata$CROPDMGEXP == "H", ]$CROPDMG * 10^2
tidydata[tidydata$CROPDMGEXP == "K", ]$CROPDMG_NOM = tidydata[tidydata$CROPDMGEXP == "K", ]$CROPDMG * 10^3
tidydata[tidydata$CROPDMGEXP == "M", ]$CROPDMG_NOM = tidydata[tidydata$CROPDMGEXP == "M", ]$CROPDMG * 10^6
tidydata[tidydata$CROPDMGEXP == "B", ]$CROPDMG_NOM = tidydata[tidydata$CROPDMGEXP == "B", ]$CROPDMG * 10^9

# Some of the event types have lower cases, convert them all to upper cases
tidydata$EVTYPE = toupper(tidydata$EVTYPE)
```

Summarize top 5 events with most damage to human health
```{r}
# summarise top 5 event causing most fatalities 
fatalitiesByType <- aggregate(cbind(FATALITIES) ~ EVTYPE, tidydata, sum, na.rm=TRUE)
fatalitiesByType <- fatalitiesByType[order(fatalitiesByType$FATALITIES, decreasing = TRUE),]
fatalitiesByType <- fatalitiesByType[1:5,]

# summarise top 5 event causing most injuries
injuriesByType <- aggregate(cbind(INJURIES) ~ EVTYPE, tidydata, sum, na.rm=TRUE)
injuriesByType <- injuriesByType[order(injuriesByType$INJURIES, decreasing = TRUE),]
injuriesByType <- injuriesByType[1:5,]
```

Summarize top 5 events with most economic damage
```{r}
# summarise top 5 events causing most crop damage
cropDmgByType <- aggregate(cbind(CROPDMG_NOM) ~ EVTYPE, tidydata, sum, na.rm=TRUE)
cropDmgByType <- cropDmgByType[order(cropDmgByType$CROPDMG_NOM, decreasing = TRUE),]
cropDmgByType <- cropDmgByType[1:5,]

# summarise top 5 events causing most property damage
propDmgByType <- aggregate(cbind(PROPDMG_NOM) ~ EVTYPE, tidydata, sum, na.rm=TRUE)
propDmgByType <- propDmgByType[order(propDmgByType$PROPDMG_NOM, decreasing = TRUE),]
propDmgByType <- propDmgByType[1:5,]
```

# Results

#### Which types of events are most harmful to population health?
Plot the data
```{r}
par(mfrow = c(1, 2), oma = c(0, 0, 2, 0))
par(mar = c(6, 4, 3, 0))
barplot(fatalitiesByType$FATALITIES, names.arg = fatalitiesByType$EVTYPE, main = "Fatalities", 
    ylab = "Fatalities", cex.axis = 0.8, cex.names = 0.7, las = 2)
barplot(injuriesByType$INJURIES, names.arg = injuriesByType$EVTYPE, main = "Injuries", 
    ylab = "Injuries", cex.axis = 0.8, cex.names = 0.7, las = 2)
title("Top 5 weather events from 1950-2011 \ncausing fatalities or injuries ", 
    outer = TRUE)
```

Tornado is the most damaging event to human life by far. It incurs the highest amount of injuries and fatalities compared to other weather events. Also note that there are a lot more injuries than fatalities across the top five weather events.



#### Which types of events have the greatest economic consequence?
Plot the data
```{r}
par(mfrow = c(1, 2), oma = c(0, 0, 2, 0))
par(mar = c(8, 4, 3, 0))
barplot(propDmgByType$PROPDMG / 10^9, names.arg = propDmgByType$EVTYPE, main = "Property damages", 
    ylab = "Billions of USD", cex.axis = 0.8, cex.names = 0.7, las = 2)
barplot(cropDmgByType$CROPDMG / 10^9, names.arg = cropDmgByType$EVTYPE, main = "Crop damages", 
    ylab = "Billions of USD", cex.axis = 0.8, cex.names = 0.7, las = 2)
title("Top 5 weather events from 1950 - 2011 \ncausing property or economic damage", 
    outer = TRUE)
```

Flood and hurricane/typhoon causes the greatest economic damage. However drought is the main contributor to crop damage. Also note that the nominal damage incurred on property are far larger than the nominal damage incurred on crops.
